04_analysis

### 2023-09-13 04_analysis.Rmd
Update 04_analysis.Rmd
added some GAM models

Bind ALL DRIFTS into one dataframe
```{r}
allDrifts <- rbind(fulltrack07, fulltrack08, fulltrack10, fulltrack12, fulltrack13, fulltrack14, 
                   fulltrack16, fulltrack18, fulltrack19, fulltrack20, fulltrack21, fulltrack22, fulltrack23)
```

@@

ggplot(avgdB, aes(x = factor(frequency, levels = unique(frequency)), y = dB)) + 
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

@@

@@

# Salinity at 400 m
ggplot(data = allDrifts, aes(sal400, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.005, color = "black")
```
### GAMs

@@

>>>>

Bind ALL DRIFTS into one dataframe
```{r}
allDrifts <- rbind(fulltrack07, fulltrack08, fulltrack10, fulltrack12, fulltrack13, fulltrack14, 
                   fulltrack16, fulltrack18, fulltrack19, fulltrack20, fulltrack21, fulltrack22, fulltrack23)
save(allDrifts, file = "allDrifts.rda")
```

@@

ggplot(avgdB, aes(x = factor(frequency, levels = unique(frequency)), y = dB)) + 
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
Looking at TOl2000 sound levels for each buoy and where Whale Detections happen
```{r fig.width=10}
ggplot(fulltrack23, aes(x = UTC, y = TOL_2000)) + geom_line() + 
  geom_smooth() + 
  geom_point(aes(color = species, size = 2)) +
  geom_point(color = shipin10km) +
  scale_color_discrete(na.translate=FALSE)
```

@@

```{r}
ggplot(allDrifts, aes(factor(BWpresence), TOL_2000)) + geom_boxplot()
```

@@

# Salinity at 400 m
ggplot(data = allDrifts, aes(sal400, fill = factor(BWpresence))) +
  geom_histogram(binwidth = 0.005, color = "black")
```

### GAMs

@@

Top 5 environmental variables (by RF) with one drift
```{r}
# Environ only
model1 <- gam(BWpresence>0 ~ depth + sst_mean + mldTemp + ttDepth + sal400,
              data = fulltrack22, family = binomial, method = "REML")

plot(model1, trans = plogis, all.terms = TRUE)
k.check(model1)
summary(model1)
gam.check(model1)
# 16% deviance explained (buoy 7)
# 0.68% deviance explained (buoy 8)
# 8.08% deviance explained (buoy 10)
# 0.64% deviance explained (buoy 12)
# 6.94 % deviance explained (buoy 13)
# 19.5% deviance explained (buoy 14)
# 12.2% deviance explained (buoy 16)
# 8.15% deviance explained (buoy 18)
# 7.7% deviance explained (buoy 19)
# 5.5% deviance explained (buoy 20)
# 33.5% deviance explained (buoy 21)
# 13.5% deviance explained (buoy 22)
# 12.1% deviance explained (buoy 23)

# Add soundscape metrics to model 1
model2 <- gam(BWpresence>0 ~ depth + sst_mean + mldTemp + ttDepth + sal400
              + TOL_125 + TOL_2000,
              data = fulltrack22, family = binomial, method = "REML")

#plot(model2, trans = plogis) ## doesn't work
summary(model2)
gam.check(model2)
# 18.3% deviance explained (buoy 7)
# 1.25% deviance explained (buoy 8)
# 8.56% deviance explained (buoy 10)
# 1.39% deviance explained (buoy 12)
# 11.1% deviance explained (buoy 13)
# 28% deviance explained (buoy 14)
# 15% deviance explained (buoy 16)
# 9.92% deviance explained (buoy 18)
# 8.57% deviance explained (buoy 19)
# 7.3% deviance explained (buoy 20)
# 38.3% deviance explained (buoy 21)
# 15.2 % deviance explained (buoy 22)
# 20.4% deviance explained (buoy 23)
```
GAM with FULL DATASET (not great)
```{r}
alldriftGAM <- gam(BWpresence>0 ~ depth + sst_mean + mldTemp + ttDepth + sal400
              + TOL_125 + TOL_2000,
              data = allDrifts, family = binomial, method = "REML")
summary(alldriftGAM)
gam.check(alldriftGAM)
```

GAMs with s()
```{r}
# Environ only, with s()
model3 <- gam(BWpresence>0 ~ s(depth) + s(sst_mean) + s(mldTemp) + 
                s(ttDepth) + s(sal400),
              data = fulltrack21, family = binomial, method = "REML")

plot(model3, trans = plogis)
summary(model3)
gam.check(model3)
# 61.9 % deviance explained (buoy 7)
# 1.42% deviance explained (buoy 8)
# 12.5% deviance explained (buoy 10)
# 1.66% deviance explained (buoy 12)
# 12.8% deviance explained (buoy 13)
# 41.4% deviance explained (buoy 14)
# 39% deviance explained (buoy 16)
# 16.6% deviance explained (buoy 18)
# 15.5% deviance explained (buoy 19)
# 5.8% deviance explained (buoy 20)
## Iteration limit reached without full convergence -- 71.7% deviance explained (buoy 21) ???
# 18.6% deviance explained (buoy 22)
# 49.6% deviance explained (buoy 23)

#Add soundscape metrics to model 3
model4 <- gam(BWpresence>0 ~ s(depth) + s(sst_mean) + s(mldTemp) + 
                s(ttDepth) + s(sal400) + s(TOL_125) + s(TOL_2000),
              data = fulltrack10, family = binomial, method = "REML")

plot(model4, trans = plogis)
summary(model4)
gam.check(model4)
# 14.5% deviance explained

# mix and match?
model5 <- gam(BWpresence>0 ~ depth + sst_mean + s(mldTemp) + 
                s(ttDepth) + sal400,
              data = fulltrack21, family = binomial, method = "REML")

summary(model5)
gam.check(model5)
# 57.4 % deviance explained
plot(model5)

```
Higher edfs describe more wiggly curves. Doesn't mean significant
